Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-05-08 paper.tex
Parens
Deal with rotations

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern \cref{fig:aniso}. This has such consequences as reduced accuracy of
computed functions due to necessity to interpolate the image during
rotation. Although we show in \cref{sec:rotqua} that even nearest-neighbor
interpolation results only in insignificant loss of quality. Rotation of the
pattern has same drawbacks related to interpolation of the sample which are most
noticeable at small correlation lengths. The both methods (rotation of the
sample or the pattern) produce nearly equivalent result if the sample is
homogeneous and an average pore size is much less than the sample's size (see
\cref{sec:rotrot})
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

\subsection{Directional and averaged computations}
When a sample is known to be isotropic, the correlation function may be
calculated multiple times, each time applying a random rotation of the sample,
and then averaged across rotations to gather more statistics. This process is
illustrated on \cref{fig:workflow}. This way we obtain similar results to
sampling procedures used in \cite{berryman1988,SMITH1988176,malmir2018}.

@@

To summarize, \code{CorrelationFunctions.jl} allows to compute 3-point
correlation functions averaged in all directions (by randomly rotating the
sample), along predefined number of directions (e.g., orthogonal and diagonal by
rotation in these directions) or along the direction of major anisotropy (that
can be determined using \code{$detect\_anisotropy$} function).  Full correlation
map for 3-point functions is impractical due its huge size and is not
implemented.

@@

We have made execuation time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et~al
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a web-like pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach based on right triangle is in
its parallelizability and ease of implementation. A drawback is that an angle
between catheti of the sampling pattern is fixed to be $\theta = \frac{\pi}{2}$,
but can be mitigated by rotation of the pattern as a whole by an equivalent
operation of rotation of a sample. These rotations can be random as in the
Malmir et al. algorithm, or it can be a single rotation to align catheti of the
sampling triangle with the direction of anisotropy (\cref{fig:aniso}). The
latter allows computing directional 3-point correlation functions for
anisotropic structures similar to 2-point counterparts
\cite{10.1063/1.4867611,EPL1}.

>>>>

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern (\cref{fig:aniso}). This has such consequences as reduced accuracy of
computed functions due to necessity to interpolate the image during
rotation. Although we show in \cref{sec:rotqua} that even nearest-neighbor
interpolation results only in insignificant loss of quality. Rotation of the
pattern has same drawbacks related to interpolation of the sample which are most
noticeable at small correlation lengths. The both methods (rotation of the
sample or the pattern) produce nearly equivalent result if the sample is
homogeneous and an average pore size is much less than the sample's size (see
\cref{sec:rotrot})
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

\subsection{Directional and averaged computations}
When a sample is known to be isotropic, the correlation function may be
calculated multiple times, each time applying a random rotation to the sample or
the pattern, and then averaged across rotations to gather more statistics. This
process is illustrated on \cref{fig:workflow}. This way we obtain similar
results to sampling procedures used in
\cite{berryman1988,SMITH1988176,malmir2018}.

@@

To summarize, \code{CorrelationFunctions.jl} allows to compute 3-point
correlation functions averaged in all directions (by randomly rotating the
sample or the pattern), along predefined number of directions (e.g., orthogonal
and diagonal by rotation in these directions) or along the direction of major
anisotropy (that can be determined using \code{$detect\_anisotropy$} function).
Full correlation map for 3-point functions is impractical due its huge size and
is not implemented.

@@

We have made execuation time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et~al
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a web-like pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach based on right triangle is in
its parallelizability and ease of implementation. A drawback is that an angle
between catheti of the sampling pattern is fixed to be $\theta = \frac{\pi}{2}$,
but it can be mitigated by choosing a random pattern or rotating the
pattern. The rotations can be random as in the Malmir et al. algorithm, or it
can be a single rotation to align catheti of the sampling triangle with the
direction of anisotropy (\cref{fig:aniso}). The latter allows computing
directional 3-point correlation functions for anisotropic structures similar to
2-point counterparts \cite{10.1063/1.4867611,EPL1}.
